NMRlipids Databank makes data-driven analysis of biomembrane properties accessible for all

### 2023-03-24 main.tex
Update pdfs

Using workflow depicted in Fig.~\ref{fig:POPCPOPEapls}D, we first calculated the cholesterol flip--flop rates from all the simulations available in the NMRlipids databank. Flip--flops were observed for cholesterol, % (634 events in 77 different simulations), 
DCHOL (18,19-di-nor-cholesterol), %, 16 flip-flops in 1 simulation) 
DOG (1,2-dioleoyl-\textit{sn}-glycerol), %, 44 flip-flops in 4 simulations) 
and SDG (1-stearoyl-2-docosahexaenoyl-\textit{sn}-glycerol). The observed cholesterol flip--flop rates, ranging between 0.001--1.6\,\textmu{}s$^{-1}$ with the mean of 0.16\,\textmu{}s$^{-1}$ and median of 0.07\,\textmu{}s$^{-1}$, are in line with the previously reported values from atomistic MD simulations~\cite{gu19,javanainen19,baral20}. The flip--flop rate of DCHOL, 0.2\,\textmu{}s$^{-1}$, was close to the average value of cholesterol, while average rates for diacylglycerols DOG and SDG were higher than for cholesterol, 0.4\,\textmu{}s$^{-1}$ and 0.5\,\textmu{}s$^{-1}$, respectively. Flip--flops were not observed for other lipids, giving the upper limits for PC lipid flip--flop rate of 9$\cdot$10$^{-6}$\,\textmu{}s$^{-1}$ and for ceramide (\textit{N}-palmitoyl-\textsc{D}-\textit{erythro}-sphingosine) of 0.002\,\textmu{}s$^{-1}$. Thus, available data in the NMRlipids databank suggests that the lipid flip--flip rate decreases in the order of diacylglycerols $>$ cholesterol $>$ other lipids including ceramides. However, the amount of data for diacylgycerols (8 simulations with Lipid17 force field) and ceramide (3 simulations with CHARMM36 force field) is less than for cholesterol (83 simulations), thus we cannot fully exclude the effect of force field or composition on this comparison.

Nevertheless, we used the workflow depicted in Fig.~\ref{fig:POPCPOPEapls}E to analyse how flip--flop rates calculated from the NMRlipids databank depend on membrane properties. Cholesterol flip--flop rates and their histograms are plotted as a function of membrane thickness, lateral density and order in Figs.~\ref{fig:flip-flops} B--D. The results reveal a non-linear correlation between cholesterol flip--flop rate and membrane packing which is depicted as area per lipid. Flip--flop rates increase by an order of magnitude when membrane packing density decreases and a major jump is observed at low membrane packing. The order of magnitude changes in cholesterol flip--flop rate with the membrane composition may have major implications in understanding lipid trafficking and membrane biochemistry~\cite{gu19,baral20}. Because the results from the NMRlipids databank are averaged over large range of membrane compositions and force fields, they show that the strong dependence of cholesterol flip--flop rate on membrane properties is not limited to certain lipid compositions or force fields used in previous studies~\cite{gu19,javanainen19,baral20}.

@@

To this end, we first calculated the water permeability through membranes from all simulations in the NMRlipids databank using the workflow depicted in Fig.~\ref{fig:POPCPOPEapls}A. The resulting non-zero values range between 0.3 and 322\,\textmu{}m/s with the mean and median of 14\,\textmu{}m/s and 8\,\textmu{}m/s, respectively. These values agree with the previously reported simulation results~\cite{venable19,camilo2022}, but are on average slightly larger than experimental values reported for PC lipids in the liquid crystalline phase,~0.19--0.33\,\textmu{}m/s~\cite{jansen95}. Using the workflow depicted in Fig.~\ref{fig:POPCPOPEapls}B, we then plotted the observed permeabilities and their histogrammed values in Figs.~\ref{fig:permeability}B--E as a function of temperature, membrane thickness, area per lipid, and acyl chain order. As expected, the permeability increases with the temperature, giving an average energy barrier of 17\,$\pm$\,4\,$k_\mathrm{B}T$ for the water permeation from the Arrhenius plot in Fig.~\ref{fig:permeability}B. On the other hand, the permeability of water decreases on average when membranes become more packed, \textit{i.e.}, with decreasing area per lipid and increasing thickness and acyl chain order (Figs.~\ref{fig:permeability} C--E). Permeation of water through bilayers depends on membrane properties also according to previous studies, but there is no established consensus on whether the area per lipid~\cite{nagle08} or bilayer thickness~\cite{frallicciardi22} is the main parameter determining the permeability. 
%is previously reported from simulations and experiments for some systems, but not for all~\cite{mathai07,venable19,frallicciardi22}. 
Our analysis over the NMRlipids databank, containing significantly more data than what was available in previous studies, suggest non-linear dependencies on both of these parameters, yet the linear correlation would be a good approximation for thicknesses above $\sim$\,3.9\,nm and area per lipids below $\sim$69\,\AA{}$^2$ (insets in Figs.~\ref{fig:permeability} C--D).
%within the range of experimentally reported values  permeability which is 1-2 orders of magnitude slower than osmotic permeability \cite{jansen95,??}. 
Clear dependencies of permeability on charged lipids, cholesterol, POPE, or hydration level were not observed in Fig.~\ref{fig:permeationSI} in the supplementary information.
%
%%Several models on how water dynamics depends on membrane properties have been proposed~\cite{mathai07,nitsche13,nitsche16,shinoda16,venable19,frallicciardi22}.

To analyze how water diffusion anisotropy depends on membrane properties in a multi-lamellar lipid bilayer system, we calculated the water diffusion parallel to the membrane surface from all simulations in the NMRlipids databank using the workflow depicted in Fig.~\ref{fig:POPCPOPEapls}A. The parallel diffusion coefficient of water, $D_{||}$, decreases with reduced hydration and increases with the temperature, but dependencies on membrane area per lipid, thickness, or fraction of charged lipids were not observed in Figs.~\ref{fig:permeability} and~\ref{fig:diffusionSI} when the workflow depicted in Fig.~\ref{fig:POPCPOPEapls}B was used. Simulation results are close to the experimental values with low hydration levels in Fig.~\ref{fig:permeability} F, but increase to approximately two times higher than the experimental value for bulk water diffusion value (3.1$\cdot$\,10$^{-9}$\,m$^2$/s at 313\,K~\cite{khakimov08}) with high hydration levels. This is not surprising as the most common water model used in membrane simulations, TIP3P, overestimates the bulk water diffusion~\cite{pathirannahalage21}. To estimate the diffusion anisotropy of water, $D_\mathrm{\perp}/D_{||}$, in multilamellar membrane system, the permeability coefficients of water through membranes were translated to perpendicular diffusion coefficients, $D_\mathrm{\perp}$, using the Tanner equation~\cite{tanner78,wasterby02}. The resulting perpendicular diffusion coefficients are approximately five orders of magnitude smaller than lateral diffusion coefficients of water (Figs.~\ref{fig:permeability} G and H), which is at the upper limit of the anisotropy estimated from the experimental data~\cite{nitsche19}. Significant increase in the diffusion anisotropy with membrane packing is observed, as $D_\mathrm{\perp}/D_{||}$ drifts away from one with decreasing area per lipid and increasing thickness in Figs.~\ref{fig:permeability} G and H. This follows from decreasing water permeability with membrane packing (Figs.~\ref{fig:permeability} C and D), while lateral diffusion remains approximately constant (Figs.~\ref{fig:diffusionSI} A and C).

@@

\begin{table}[t]
    \centering
    \begin{tabular}{p{5.0cm}  p{5.0cm}  p{4.0cm}}
    Type of application     & Practical examples & Target group \\
    \hline
    Analyses of rare phenomena               & Lipid flip--flops, water permeation (Figs.~\ref{fig:flip-flops} and~\ref{fig:permeability}) & Membrane scientists \\
    \\
    Correlations between membrane properties & 
    Membrane structural properties, water dynamics (Figs.~\ref{fig:quality} and~\ref{fig:permeability}) & 
    Membrane scientists \\
    \\
    Applications that are outside typical scope of MD simulations & 
    Anisotropic water diffusion for pharmacokinetics and MRI imaging applications (Fig.~\ref{fig:permeability}) & 
    Scientist in fields where MD simulations are not usually applied \\
    \\
    Selection of the best simulation model for a specific application & 
    Best model for membranes with PC and PE lipids (Fig.~\ref{fig:quality}), lipid headgroup conformations~\cite{bacle21}, 
    packing of PS~\cite{antila22b} and PE (Fig.~\ref{fig:POPCPOPEapls}) containing membranes. &
    Scientists using MD simulations \\
    \\
    Guidance for force field development & 
    Improvements in ion binding to lipids~\cite{melcr18,melcr20} and lipid headgroup conformational ensembles~\cite{yu21,dickson22,grote20} &
    Scientists developing parameters for MD simulations \\
    \\
    Training and target data for the development of coarse grained models & 
    Optimizing parameters of coarse grained models against NMRlipids databank, extracting continuum parameters for membranes. &
    Scientists developing and using coarse grained MD simulations \\
    \\
    Training set for machine learning applicatons &
    Programmatic access to the data and results enables training of machine learning type of models for various applications, such as predictions of membrane properties from composition & Scientists building and using machine learning applications for biomolecules.  
    \end{tabular}
    \caption{Examples on applications of the NMRlipids databank.}
    \label{tab:applications}
\end{table}

>>>>

Using workflow depicted in Fig.~\ref{fig:overlay}D, we first calculated the cholesterol flip--flop rates from all the simulations available in the NMRlipids databank. Flip--flops were observed for cholesterol, % (634 events in 77 different simulations), 
DCHOL (18,19-di-nor-cholesterol), %, 16 flip-flops in 1 simulation) 
DOG (1,2-dioleoyl-\textit{sn}-glycerol), %, 44 flip-flops in 4 simulations) 
and SDG (1-stearoyl-2-docosahexaenoyl-\textit{sn}-glycerol). The observed cholesterol flip--flop rates, ranging between 0.001--1.6\,\textmu{}s$^{-1}$ with the mean of 0.16\,\textmu{}s$^{-1}$ and median of 0.07\,\textmu{}s$^{-1}$, are in line with the previously reported values from atomistic MD simulations~\cite{gu19,javanainen19,baral20}. The flip--flop rate of DCHOL, 0.2\,\textmu{}s$^{-1}$, was close to the average value of cholesterol, while average rates for diacylglycerols DOG and SDG were higher than for cholesterol, 0.4\,\textmu{}s$^{-1}$ and 0.5\,\textmu{}s$^{-1}$, respectively. Flip--flops were not observed for other lipids, giving the upper limits for PC lipid flip--flop rate of 9$\cdot$10$^{-6}$\,\textmu{}s$^{-1}$ and for ceramide (\textit{N}-palmitoyl-\textsc{D}-\textit{erythro}-sphingosine) of 0.002\,\textmu{}s$^{-1}$. Thus, available data in the NMRlipids databank suggests that the lipid flip--flip rate decreases in the order of diacylglycerols $>$ cholesterol $>$ other lipids including ceramides. However, the amount of data for diacylgycerols (8 simulations with Lipid17 force field) and ceramide (3 simulations with CHARMM36 force field) is less than for cholesterol (83 simulations), thus we cannot fully exclude the effect of force field or composition on this comparison.

Nevertheless, we used the workflow depicted in Fig.~\ref{fig:overlay}E to analyse how flip--flop rates calculated from the NMRlipids databank depend on membrane properties. Cholesterol flip--flop rates and their histograms are plotted as a function of membrane thickness, lateral density and order in Figs.~\ref{fig:flip-flops} B--D. The results reveal a non-linear correlation between cholesterol flip--flop rate and membrane packing which is depicted as area per lipid. Flip--flop rates increase by an order of magnitude when membrane packing density decreases and a major jump is observed at low membrane packing. The order of magnitude changes in cholesterol flip--flop rate with the membrane composition may have major implications in understanding lipid trafficking and membrane biochemistry~\cite{gu19,baral20}. Because the results from the NMRlipids databank are averaged over large range of membrane compositions and force fields, they show that the strong dependence of cholesterol flip--flop rate on membrane properties is not limited to certain lipid compositions or force fields used in previous studies~\cite{gu19,javanainen19,baral20}.

@@

To this end, we first calculated the water permeability through membranes from all simulations in the NMRlipids databank using the workflow depicted in Fig.~\ref{fig:overlay}D. The resulting non-zero values range between 0.3 and 322\,\textmu{}m/s with the mean and median of 14\,\textmu{}m/s and 8\,\textmu{}m/s, respectively. These values agree with the previously reported simulation results~\cite{venable19,camilo2022}, but are on average slightly larger than experimental values reported for PC lipids in the liquid crystalline phase,~0.19--0.33\,\textmu{}m/s~\cite{jansen95}. Using the workflow depicted in Fig.~\ref{fig:overlay}E, we then plotted the observed permeabilities and their histogrammed values in Figs.~\ref{fig:permeability}B--E as a function of temperature, membrane thickness, area per lipid, and acyl chain order. As expected, the permeability increases with the temperature, giving an average energy barrier of 17\,$\pm$\,4\,$k_\mathrm{B}T$ for the water permeation from the Arrhenius plot in Fig.~\ref{fig:permeability}B. On the other hand, the permeability of water decreases on average when membranes become more packed, \textit{i.e.}, with decreasing area per lipid and increasing thickness and acyl chain order (Figs.~\ref{fig:permeability} C--E). Permeation of water through bilayers depends on membrane properties also according to previous studies, but there is no established consensus on whether the area per lipid~\cite{nagle08} or bilayer thickness~\cite{frallicciardi22} is the main parameter determining the permeability. 
%is previously reported from simulations and experiments for some systems, but not for all~\cite{mathai07,venable19,frallicciardi22}. 
Our analysis over the NMRlipids databank, containing significantly more data than what was available in previous studies, suggest non-linear dependencies on both of these parameters, yet the linear correlation would be a good approximation for thicknesses above $\sim$\,3.9\,nm and area per lipids below $\sim$69\,\AA{}$^2$ (insets in Figs.~\ref{fig:permeability} C--D).
%within the range of experimentally reported values  permeability which is 1-2 orders of magnitude slower than osmotic permeability \cite{jansen95,??}. 
Clear dependencies of permeability on charged lipids, cholesterol, POPE, or hydration level were not observed in Fig.~\ref{fig:permeationSI} in the supplementary information.
%
%%Several models on how water dynamics depends on membrane properties have been proposed~\cite{mathai07,nitsche13,nitsche16,shinoda16,venable19,frallicciardi22}.

To analyze how water diffusion anisotropy depends on membrane properties in a multi-lamellar lipid bilayer system, we calculated the water diffusion parallel to the membrane surface from all simulations in the NMRlipids databank using the workflow depicted in Fig.~\ref{fig:overlay}D. The parallel diffusion coefficient of water, $D_{||}$, decreases with reduced hydration and increases with the temperature, but dependencies on membrane area per lipid, thickness, or fraction of charged lipids were not observed in Figs.~\ref{fig:permeability} and~\ref{fig:diffusionSI} when the workflow depicted in Fig.~\ref{fig:overlay}E was used. Simulation results are close to the experimental values with low hydration levels in Fig.~\ref{fig:permeability} F, but increase to approximately two times higher than the experimental value for bulk water diffusion value (3.1$\cdot$\,10$^{-9}$\,m$^2$/s at 313\,K~\cite{khakimov08}) with high hydration levels. This is not surprising as the most common water model used in membrane simulations, TIP3P, overestimates the bulk water diffusion~\cite{pathirannahalage21}. To estimate the diffusion anisotropy of water, $D_\mathrm{\perp}/D_{||}$, in multilamellar membrane system, the permeability coefficients of water through membranes were translated to perpendicular diffusion coefficients, $D_\mathrm{\perp}$, using the Tanner equation~\cite{tanner78,wasterby02}. The resulting perpendicular diffusion coefficients are approximately five orders of magnitude smaller than lateral diffusion coefficients of water (Figs.~\ref{fig:permeability} G and H), which is at the upper limit of the anisotropy estimated from the experimental data~\cite{nitsche19}. Significant increase in the diffusion anisotropy with membrane packing is observed, as $D_\mathrm{\perp}/D_{||}$ drifts away from one with decreasing area per lipid and increasing thickness in Figs.~\ref{fig:permeability} G and H. This follows from decreasing water permeability with membrane packing (Figs.~\ref{fig:permeability} C and D), while lateral diffusion remains approximately constant (Figs.~\ref{fig:diffusionSI} A and C).

@@

\begin{table}[t]
    \centering
    \begin{tabular}{p{5.0cm}  p{5.0cm}  p{4.0cm}}
    Type of application     & Practical examples & Target group \\
    \hline
    Analyses of rare phenomena               & Lipid flip--flops, water permeation (Figs.~\ref{fig:flip-flops} and~\ref{fig:permeability}) & Membrane scientists \\
    \\
    Correlations between membrane properties & 
    Membrane structural properties, water dynamics (Figs.~\ref{fig:quality} and~\ref{fig:permeability}) & 
    Membrane scientists \\
    \\
    Applications that are outside typical scope of MD simulations & 
    Anisotropic water diffusion for pharmacokinetics and MRI imaging applications (Fig.~\ref{fig:permeability}) & 
    Scientist in fields where MD simulations are not usually applied \\
    \\
    Selection of the best simulation model for a specific application & 
    Best model for membranes with PC and PE lipids (Fig.~\ref{fig:quality}), lipid headgroup conformations~\cite{bacle21}, 
    packing of PS~\cite{antila22b} and PE (Fig.~\ref{fig:quality}) containing membranes. &
    Scientists using MD simulations \\
    \\
    Guidance for force field development & 
    Improvements in ion binding to lipids~\cite{melcr18,melcr20} and lipid headgroup conformational ensembles~\cite{yu21,dickson22,grote20} &
    Scientists developing parameters for MD simulations \\
    \\
    Training and target data for the development of coarse grained models & 
    Optimizing parameters of coarse grained models against NMRlipids databank, extracting continuum parameters for membranes. &
    Scientists developing and using coarse grained MD simulations \\
    \\
    Training set for machine learning applicatons &
    Programmatic access to the data and results enables training of machine learning type of models for various applications, such as predictions of membrane properties from composition & Scientists building and using machine learning applications for biomolecules.  
    \end{tabular}
    \caption{Examples on applications of the NMRlipids databank.}
    \label{tab:applications}
\end{table}
